```{r}
install.packages("readr")
```
```{r}
install.packages('textshaping')
```

```{r}
install.packages('ragg')
```

```{r}
install.packages('tidyverse')
```

```{r}
library(readr)
library(tidyverse)
```
```{r}
#import and display the data
getwd()
data <- read.csv('data/cema_internship_task_2023.csv', header=TRUE, sep=",")
data
```
Question 1
Is there a correlation between the number of children dewormed and the prevalence of acute malnutrition at the county level over the study period?
Since, acute malnutrition data has missing values I will fill them with a mean value and with 0, and see which value gives a resonable answer.
```{r}
#return columns with missing data
na_columns <- colnames(data)[apply(data, 2, anyNA)]
na_columns
```
```{r}
#create mean value ignoring the missing values
mean_acute <- mean(data$Acute.Malnutrition, na.rm=TRUE)
mean_acute
```
```{r}
#create new column with mean values using mutate
data_replace_acute <- data %>% mutate(replace_acute = ifelse(is.na(Acute.Malnutrition), mean_acute, Acute.Malnutrition))
data_replace_acute
```
Calculating the correlation
```{r}
install.packages('lme4')
install.packages('pbkrtest')
install.packages('car')
install.packages("rstatix")

install.packages("ggpubr", repos = "https://cloud.r-project.org/", dependencies = TRUE)


```
```{r}
install.packages("nloptr")
```
```{r}
#The Spearman correlation method computes the correlation between the rank of x and the rank of y variable
#Correlation between children dewormed and acute malnutrition
data_replace_acute <- data %>% mutate(replace_acute = ifelse(is.na(Acute.Malnutrition), mean_acute, Acute.Malnutrition))
data_replace_acute
```


```{r}
correlation <- cor(data_replace_acute$Total.Dewormed, data_replace_acute$replace_acute, use = "complete.obs")
cat("Correlation between Total Dewormed and Acute Malnutrition:", correlation, "\n")
```
Visually present the correlation
```{r}
# Plot the scatter plot
ggplot(data_replace_acute, aes(x = Total.Dewormed, y = replace_acute)) +
  geom_point() +
  labs(title = "Scatter Plot: Total Dewormed vs. Acute Malnutrition",
       x = "Total Dewormed",
       y = "Acute Malnutrition") +
  theme_minimal()
```
Based on the scatter plot there's no strong correlation between the dewormed children and malnutrition.
Q2.Is there a trend in the number of diarrhoea cases among children under the 5 years in the different counties?
```{r}
#diarrhoea cases per county per period
data_diarrhoea <- data %>% select(period, county, diarrhoea.cases)
data_diarrhoea
```

```{r}
#install.packages('plotly')
#install.packages("viridis")
library(viridis)
library(plotly)

# Create the interactive heatmap using plot_ly
heatmap <- plot_ly(data_diarrhoea, x = ~period, y = ~county, z = ~diarrhoea.cases,
                   type = "heatmap", colors = "viridis",
                   colorscale = "viridis", showscale = TRUE) %>%
  layout(title = "Interactive Heatmap: Diarrhoea Cases among Children Under 5",
         xaxis = list(title = "Period"),
         yaxis = list(title = "County"),
         grid = list(rows = unique(data_diarrhoea$county)),
         margin = list(l = 80, b = 80))

# Display the interactive heatmap
heatmap
```

The heatmap shows that Nairobi county tends to experience cases of diarrhoea over the period the data was collected with less cases being reported in the Aug-22, Jan-21, Jul-22 and  Sep-22. High cases we reported in Feb, Mar and April 2023.

Q3. Which is the county with highest number of dewormed children?
```{r}
# group the data by county and period
data_grouped_by_county_period <- data %>% 
  group_by(county, period)
```
```{r}
library(dplyr)

# Group the data by County and calculate average Total Dewormed and Acute Malnutrition
data_grouped_avg <- data_replace_acute %>% 
  group_by(county) %>%
  summarise(avg_Total_Dewormed = mean(Total.Dewormed, na.rm = TRUE),
            avg_Acute_Malnutrition = mean(replace_acute, na.rm = TRUE))
```

```{r}
# Visualize average Total Dewormed by County
ggplot(data_grouped_avg, aes(x = county, y = avg_Total_Dewormed)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(title = "Average Total Dewormed by County",
       x = "County", y = "Average Total Dewormed") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```
Nairobi has the highest numer of dewormed children

Q4. Which county has the highest number of malnutritioned kids?
```{r}
# Visualize average Acute Malnutrition by County
ggplot(data_grouped_avg, aes(x = county, y = avg_Acute_Malnutrition)) +
  geom_bar(stat = "identity", fill = "lightgreen") +
  labs(title = "Average Acute Malnutrition by County",
       x = "County", y = "Average Acute Malnutrition") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```
Q6.What is the rate of acute malnutrition, stunting and underweight for each county?
```{r}
#replace missing values the columns "stunted.6.23.months"  "stunted.0..6.months"  "stunted.24.59.months"
# Calculate the mean of stunted.6.23.months
mean_stunted_6_23 <- mean(data$stunted.6.23.months, na.rm = TRUE)


# Replace missing values with the mean for stunted.6.23.months
data_replace <- data %>% mutate(stunted.6.23.months = ifelse(is.na(stunted.6.23.months), mean_stunted_6_23, stunted.6.23.months))

# Calculate the mean of stunted.0..6.months
mean_stunted_0_6 <- mean(data$stunted.0..6.months, na.rm = TRUE)

# Replace missing values with the mean for stunted.0..6.months
data_replace <- data %>% mutate(stunted.0..6.months = ifelse(is.na(stunted.0..6.months), mean_stunted_0_6, stunted.0..6.months))

# Calculate the mean of stunted.24.59.months
mean_stunted_24_59 <- mean(data$stunted.24.59.months, na.rm = TRUE)

# Replace missing values with the mean for stunted.24.59.months
data_replace <- data %>% mutate(stunted.24.59.months = ifelse(is.na(stunted.24.59.months), mean_stunted_24_59, stunted.24.59.months))

# Replace missing values with the mean for acute malnutrition
mean_acute <- mean(data$Acute.Malnutrition, na.rm=TRUE)

#create new column with mean values using mutate
data_replace <- data %>% mutate(Acute.Malnutrition = ifelse(is.na(Acute.Malnutrition), mean_acute, Acute.Malnutrition))


data_replace
```
```{r}
#return columns with missing data
na_columns <- colnames(data_replace)[apply(data_replace, 2, anyNA)]
na_columns
```

```{r}
# Create the grouped bar chart
ggplot(data_replace, aes(x = county)) +
  geom_bar(aes(y = stunted.6.23.months, fill = "6-23 months"), position = "dodge", stat = "identity", width = 0.5) +
  geom_bar(aes(y = stunted.0..6.months, fill = "0-6 months"), position = "dodge", stat = "identity", width = 0.5) +
  geom_bar(aes(y = stunted.24.59.months, fill = "24-59 months"), position = "dodge", stat = "identity", width = 0.5) +
  labs(title = "Rates of Stunting by Age Group and County",
       x = "County", y = "Prevalence Rate (%)") +
  scale_fill_manual(name = "Age Group",
                    values = c("6-23 months" = "blue", "0-6 months" = "red", "24-59 months" = "green"),
                    labels = c("6-23 months" = "6-23 months", "0-6 months" = "0-6 months", "24-59 months" = "24-59 months")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  coord_flip()


```
The data shows for age group 0-6 months  Nairobi has the highest rate of stunted growth, for 24-59 months Garissa has the highest and 6 - 23 months Kitui has the highest.
